---
title: "Working with ATC codes"
graphics: yes
output:
  BiocStyle::html_document:
    toc_depth: 2
  BiocStyle::pdf_document:
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Working with ATC codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{atc}
  %\VignettePackage{atc}
  %\VignetteKeywords{AnnotationData,Homo_sapiens}
---

**Authors**: `r packageDescription("atc")$Author`<br />
**Modified**: 31 May, 2016<br />
**Compiled**: `r date()`

# Introduction

This package provides the Anatomical Therapeutical Chemical (ATC) Classification
data and functions that facilitate working with these. The ATC data is defined
by the WHO Collaborating Centre (WHOCC) for Drug Statistics Methodology (Oslo,
Norway), which holds also the copyright for the data and dos not allow
distribution for commercial purposes.

This packages provides two ATC databases, bound to the variables `atc` and
`atc_wido`.  The `AtcDb` object bound to the `atc` variable access a database
containing the *official* ATC data from the WHOCC web page. The `AtcDb` object
bound to `atc_wido` provides access to a database that was compiled manually
from data files from the *Wissenschaftliches Institut der AOK*
(<http://www.wido.de>), and from the online resources <http://atccodes.com> and
Wikipedia. The two databases might differ partially in their content (e.g. the
`atc_wido` contains also homeopathic compounds).

The ATC Classification system is used for the classification of active
ingredients of drugs according to the organ or system on which they act and
their therapeutic, pharmacological and chemical properties. In this system,
drugs are classified into groups at 5 different levels:

-   **First level**: indicates the anatomical main group and consists of one letter.

-   **Second level**: indicates the therapeutic main group and consists of two
    digits.

-   **Third level**: indicates the therapeutic/pharmacological subgroup and consists
    of one letter.

-   **Fourth level**: indicates the chemical/therapeutic/pharmacological subgroup
    and consists of one letter.

-   **Fifth level**: indicates the chemical substance and consists of two
    digits. For the 5th level entries also defined daily dosages (ddd) may be
    provided.

Note that national versions of the ATC classification exist with eventual
additional codes and ddds.

# Basic usage

The data structure of the database is very simple, and it is thus straight
forward to retrieve data from it.

The `columns` (or `listColumns`) or the `listTables` methods can be used to get some
information of the available values (see Database layout section below for a
more detailed description of the individual columns).

```{r }
library(atc)

## First of all print some basic information
atc

## What tables are available?
listTables(atc)

## All available columns.
listColumns(atc)
```

The `keys` method can be used to retrieve the ATC codes in the database, with
the additional argument `level` we can restrict to ATC codes from a certain
level or levels.

```{r }
## All ATC codes
head(keys(atc))

## Only ATC codes from level 1 and 3
head(keys(atc, level=c(1, 3)))
```

The `atcs` method can be used to extract additional information from the database.

```{r }
## Get the full data
head(atcs(atc))

## To retrive the full data we could also use the as.data.frame method
head(as.data.frame(atc))
```

The arguments `columns` and `filters` allow to define which columns respectively
entries (rows) should be retrieved from the database. The `atc` package implements
the same filtering framework than Bioconductor's `ensembldb` package, i.e. a
single or a `list` of filter objects can be passed to the `atcs` method to query for
specific entries.

The filters that are presently supported by the `atc` package are:

-   `AdminrouteFilter`: filter results based on the *administration route* (e.g. `O` for
    *oral*).
-   `KeyFilter`: filter results using their key values.
-   `LevelFilter`: filter results based on the ATC level.
-   `NameFilter`: filter results based on the entries' names.

In the code example below we simply query the database to get
all level 1 entries from the database.

```{r }
## Get all level 1 entries and restrict to some columns
atcs(atc, filter=LevelFilter(1), columns=c("key", "name", "level"))
```

In the example below we combine filters to retrieve all orally taken
glucocorticoids.

```{r }
## Use a KeyFilter to search for all entries that start with H02AB
## and combine that with administration route being 'O'
atcs(atc, filter=list(KeyFilter("H02AB%", condition="like"),
		      AdminrouteFilter("O")))
```

# Extract ATC codes from the official WHOCC web page

The official WHOCC web page provides a simple search functionality to retrieve
data for a user specified ATC code. The `crunchWHOCC` function we use below
recursively retrieves ATC data from this web page starting from either a
user-specified ATC code, or if not provided, for all level 1 codes and above.

```{r eval=FALSE}
## Retrieve all data from the WHOCC web page.
whoccAtc <- crunchWHOCC()
```

The function returns 3 `data.frames` that can be used to build a RSQLite
database such as provided by the `atc` package.

```{r eval=FALSE}
library(RSQLite)

## Connect to the database.
con <- dbConnect(dbDriver("SQLite"), dbname="atc.sqlite")
## Store the 3 tables.
dbWriteTable(con, name="atc", whoccAtc$atc, row.names=FALSE)
dbWriteTable(con, name="ddd", whoccAtc$ddd, row.names=FALSE)
dbWriteTable(con, name="metadata", whoccAtc$metadata, row.names=FALSE)
## Create indices for faster access.
dbGetQuery(con, "create index key_idx on atc (key);")
dbGetQuery(con, "create index level_idx on atc (level);")
dbGetQuery(con, "create index ddd_key_idx on ddd (key);")
dbDisconnect(con)
```

A `AtcDb` object can then be created using.

```{r eval=FALSE}
atcDb <- AtcDb("atc.sqlite")
atcDb
```

# Database layout

The database consists of 3 tables, `atc`, `ddd` and `metadata`. The columns for
the individual tables are listed below.

-   **atc**: the *main* table providing the mapping between the ATC codes (also
    called keys) and their names.
    -   `key`: column containing the ATC codes. The `KeyFilter` can be used to filter on
        this column.
    -   `name`: column with the name associated with the key. The `NameFilter` can be
        used to filter on this column.
    -   `level`: the level of the ATC code (from 1 to 5). The `LevelFilter` can be used
        to filter on this column.

-   **ddd**: the table providing the *defined daily dosage* (ddd) for the ATC
    entries of level 5. Thus, this table contains only values for codes of level
    5, but not for all of them. Note also, that this table can contain multiple
    dosages for the same ATC code.
    -   `key`: the ATC code.
    -   `ddd`: the defined daily dose.
    -   `unit`: the unit of the dose.
    -   `administration_route`: the route of administration of the drug. The
        `AdminrouteFilter` can be used to filter on this column.
    -   `note`: optional additional information.

-   **metadata**: table providing additional information (creation time, source
    etc). Values are stored as key-value pairs.
    -   `name`: the name of the key.
    -   `value`: its value.
